---
title: "Practical: Network meta-analysis"
output: 
  pdf_document
format:
  pdf
links: 
  - name: ""
    icon: slideshare
    icon_pack: fab
    # url: /slides/07_NMA/index.html
  - name: "PDF version"
    # url: practical6.pdf
    icon: file-pdf
    icon_pack: fas
---

<!-- ::: {.content-hidden when-format="pdf"} -->
<!-- {{< include /practical/_info.qmd >}} -->
<!-- ::: -->


## 1. Introduction

The data from the smoking cessation studies discussed in the lecture are
included in the file [`smoke.Rdata`](smoke.Rdata), which includes a list with the
relevant variables. The data include the following variables:

-   `NS`: the total number of studies included in our analysis (24);

-   `NT`: the total number of interventions considered (4);

-   `na`: a vector containing the number of arms included in each of the
    studies;

-   `r`: a matrix with `NS` rows and `NT` columns, containing the number
    of subjects that in each study and under each treatment arms have
    been observed to quit smoking;

-   `n`: a matrix with `NS` rows and `NT` columns, containing the total
    number of subjects observed in each study and under each treatment
    arms;

-   `t`: a matrix with `NS` rows and 3 columns, identifying the label
    associated with the treatments included in each of the studies.
    Notice that there are 3 columns because all studies have at most 3
    treatment involved (i.e. all studies compare either 2 or 3
    treatments --- cfr. the lecture slides). The treatments are labelled
    as 1 = No intervention; 2 = Self-help; 3 = Individual counselling; 4
    = Group counselling.

The <tt>R</tt> script [`NMA.R`](NMA.R) guides you through the
analysis.

## 2. Fixed effects NMA

1.  Save the `.Rdata` file to your computer and then load the data into
    your <tt>R</tt> workspace. Explore the data list
    and check you understand what the data mean.

2.  Save the file [`smokefix_model.txt`](smokefix_model.txt) to your computer. Open it and go
    through the code, making sure you understand it.

3.  Follow the <tt>R</tt> script and run the model
    calling <tt>OpenBUGS</tt> in the background,
    firstly without burn-in iterations.

4.  Follow the <tt>R</tt> script and produce
    traceplots to check convergence.

5.  Follow the <tt>R</tt> script and re-run the model,
    this time monitoring all the relevant parameters.

## 3. Random effects NMA

1.  Save the file [`smokere_model.txt`](smokere_model.txt) to your computer. Open it and go
    through the code, making sure you understand it. Inspect
    particularly the difference with the fixed effects model.

2.  Follow the <tt>R</tt> script and run the model
    calling <tt>OpenBUGS</tt> in the background.
    Comment on convergence and compare the output with the fixed
    effect model. Are there any striking differences?

3.  Follow the <tt>R</tt> script and perform the
    economic analysis, combining the output of your Bayesian Random
    Effects NMA and using `BCEA` to post-process the results.


## 4. Visualising heterogeneity and "random" vs "fixed" effects

The R script [`make_plots.R`](make_plots.R) can be used to post-process the data and the outputs from the two models above and produce the visualisations shown in the lecture slides. 

The script uses the current version of the package `bmhe`, which includes all the utility functions helpful for the various practicals in the module. You need to install it from `GitHub`, using the following commands.

```{r echo=TRUE,eval=FALSE}
# First installs the package 'remotes', which can be used to install packages directly from GitHub
install.packages("remotes")

# Then use it to install 'bmhe'
remotes::install_github("giabaio/bmhe_utils")
```

The code assumes you have downloaded the data and model codes referred above. If you've actually run the models in points 2. and 3. you don't need to run the following lines

```{r eval=FALSE}
# FIXED EFFECTS MODEL
# Initial values
inits <- list(list(mu=rep(0,24), d=c(NA,0,0,0)),
              list(mu=rep(-1,24), d=c(NA,1,1,1)))
res <- bugs(model="smokefix_model.txt", data=smoke.list, inits=inits,
           parameters=c("d","or","L","pq"),
           n.chains=2, n.burnin=1000, n.iter=20000)
```
and
```{r eval=FALSE}
# RANDOM EFFECTS MODEL.  Check convergence of random effects SD in particular
inits <- list(list(mu=rep(0,24), d=c(NA,0,0,0), sd=1),
             list(mu=rep(-1,24), d=c(NA,1,1,1), sd=2))

res2 <- bugs(model="smokere_model.txt", data=smoke.list, inits=inits,
           parameters=c("or", "d", "sd", "pq", "L"),
           n.chains=2, n.burnin=1000, n.iter=20000)
```

from the file `make_plots.R`: if you've already run the script `NMA.R`, then the objects `res` and `res2` will already be present in your `R` workspace.
